Why is it that statistical tests for residuals are not widely used? An explanation using visual inference.

### 2023-03-08 paper_comparison/paper_comparison/paper_comparison.Rmd
finish KL formula in appendix
effect size explanation

Effect size $E$ can be defined as the difference of a parameter for a particular model or distribution, or a statistic derived from a sample. Importantly, it needs to reflect the treatment we try to measure. According to the underlying simulated model, we define the effect size

$$E = \frac{1}{2}\left(\boldsymbol{\mu}_z'[diag(\boldsymbol{R}_a\sigma^2)]^{-1}\boldsymbol{\mu}_z\right)$$

for the non-linearity model, and

$$E = \frac{1}{2}\left(log\frac{|diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')|}{|diag(\boldsymbol{R}_a)|} - n + tr(diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')^{-1}diag(\boldsymbol{R}_a))\right)$$
for the heteroskedasticity model, where $diag(.)$ is the diagonal matrix constructed from the diagonal elements of a matrix, $\sigma^2\boldsymbol{I}$ is the assumed covariance matrix of the error term when $H_0$ is true, $\boldsymbol{V}$ is the actual covariance matrix of the error term, $\boldsymbol{R}_a = \boldsymbol{I}_n - \boldsymbol{H}_a$ is the residual operator, $\boldsymbol{H}_a = \boldsymbol{X}_a(\boldsymbol{X}_a'\boldsymbol{X}_a)^{-1}\boldsymbol{X}_a'$ is the hat matrix, $\boldsymbol{X}_a = (\boldsymbol{1},  \boldsymbol{X})$ is the matrix of predictors including the intercept used in the regression equation, and $\boldsymbol{\mu}_z = \boldsymbol{R}_a\boldsymbol{Z}\boldsymbol{\beta}_z$ is the expected values of residuals with $\boldsymbol{Z}$ be any higher order terms of $\boldsymbol{X}$ leave out by the regression equation and $\boldsymbol{\beta}_z$ be the corresponding coefficients.

Both equations above are derived from the Kullback-Leibler divergence (citation here?) and the details are provided in the Appendix.

>>>>

Effect size in statistics measures the strength of the signal relative to the noise. It is surprisingly difficult to quantify in general, even for simulated data as used in this experiment.

For the non-linearity model, the key items defining effect size are sample size ($n$) and variance of the error term ($\sigma^2$), and so effect size would be roughly calculated as $\frac{\sqrt(n)}{\sigma}$. As sample size increases the effect size would increase, but as variance increases the effect size decreases. However, it is not clear how the additional parameter for the model polynomial order, $k$, should be incorporated. Intuitively, the large $k$ means more complex pattern, which likely means effect size would decrease. For the purposes of our calculations we have chosen to use an approach based on Kullback-Leibler divergence [@kullback1951information], coupled with simulation. This formulation defines effect size to be:

$$E = \frac{1}{2}\left(\boldsymbol{\mu}_z'(diag(\boldsymbol{R}_a\sigma^2))^{-1}\boldsymbol{\mu}_z\right)$$

\noindent where $\boldsymbol{\mu}_z = \boldsymbol{R}_a\boldsymbol{Z}\boldsymbol{\beta}_z$ is the expected values of residuals with $\boldsymbol{Z}$, $\boldsymbol{R}_a = \boldsymbol{I}_n - \boldsymbol{H}_a$ is the residual operator, $\boldsymbol{H}_a = \boldsymbol{X}_a(\boldsymbol{X}_a'\boldsymbol{X}_a)^{-1}\boldsymbol{X}_a'$ is the hat matrix. For the non-linearity model this can be rewritten as

$$E = \frac{1}{2}\left(log\frac{|diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')|}{|diag(\boldsymbol{R}_a)|} - n + tr(diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')^{-1}diag(\boldsymbol{R}_a))\right)$$

\noindent where $\boldsymbol{V}$ is the covariance matrix of the error term.

<!-- Effect size $E$ can be defined as the difference of a parameter for a particular model or distribution, or a statistic derived from a sample. Importantly, it needs to reflect the treatment we try to measure. According to the underlying simulated model, we define the effect size

$$E = \frac{1}{2}\left(\boldsymbol{\mu}_z'(diag(\boldsymbol{R}_a\sigma^2))^{-1}\boldsymbol{\mu}_z\right)$$

for the non-linearity model, and -->

In the heteroskedasticity model, the key elements for measuring effect size are sample size, $n$, and the ratio of the biggest variance to smallest variance, $b$. Larger values of both would produce higher effect size. However, it is not clear how to incorporate the additional shape parameter, $a$. Thus the same approach is used here, where the formula can be written as:

$$E = \frac{1}{2}\left(log\frac{|diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')|}{|diag(\boldsymbol{R}_a)|} - n + tr(diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')^{-1}diag(\boldsymbol{R}_a))\right)$$
\noindent where $diag(.)$ is the diagonal matrix constructed from the diagonal elements of a matrix. <!--, $\sigma^2\boldsymbol{I}$ is the assumed covariance matrix of the error term when $H_0$ is true, $\boldsymbol{V}$ is the actual covariance matrix of the error term, $\boldsymbol{R}_a = \boldsymbol{I}_n - \boldsymbol{H}_a$ is the residual operator, $\boldsymbol{H}_a = \boldsymbol{X}_a(\boldsymbol{X}_a'\boldsymbol{X}_a)^{-1}\boldsymbol{X}_a'$ is the hat matrix, $\boldsymbol{X}_a = (\boldsymbol{1},  \boldsymbol{X})$ is the matrix of predictors including the intercept used in the regression equation, and $\boldsymbol{\mu}_z = \boldsymbol{R}_a\boldsymbol{Z}\boldsymbol{\beta}_z$ is the expected values of residuals with $\boldsymbol{Z}$ be any higher order terms of $\boldsymbol{X}$ leave out by the regression equation and $\boldsymbol{\beta}_z$ be the corresponding coefficients. -->

Derivations for these equations are provided in the Appendix.

To compute the effect size for each lineup we simulate ??

### 2023-03-08 paper_comparison/paper_comparison/paper_comparison.tex
effect size explanation

Effect size \(E\) can be defined as the difference of a parameter for a
particular model or distribution, or a statistic derived from a sample.
Importantly, it needs to reflect the treatment we try to measure.
According to the underlying simulated model, we define the effect size

\[E = \frac{1}{2}\left(\boldsymbol{\mu}_z'[diag(\boldsymbol{R}_a\sigma^2)]^{-1}\boldsymbol{\mu}_z\right)\]

for the non-linearity model, and

\[E = \frac{1}{2}\left(log\frac{|diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')|}{|diag(\boldsymbol{R}_a)|} - n + tr(diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')^{-1}diag(\boldsymbol{R}_a))\right)\]
for the heteroskedasticity model, where \(diag(.)\) is the diagonal
matrix constructed from the diagonal elements of a matrix,
\(\sigma^2\boldsymbol{I}\) is the assumed covariance matrix of the error
term when \(H_0\) is true, \(\boldsymbol{V}\) is the actual covariance
matrix of the error term,
\(\boldsymbol{R}_a = \boldsymbol{I}_n - \boldsymbol{H}_a\) is the
residual operator,
\(\boldsymbol{H}_a = \boldsymbol{X}_a(\boldsymbol{X}_a'\boldsymbol{X}_a)^{-1}\boldsymbol{X}_a'\)
is the hat matrix,
\(\boldsymbol{X}_a = (\boldsymbol{1}, \boldsymbol{X})\) is the matrix of
predictors including the intercept used in the regression equation, and
\(\boldsymbol{\mu}_z = \boldsymbol{R}_a\boldsymbol{Z}\boldsymbol{\beta}_z\)
is the expected values of residuals with \(\boldsymbol{Z}\) be any
higher order terms of \(\boldsymbol{X}\) leave out by the regression
equation and \(\boldsymbol{\beta}_z\) be the corresponding coefficients.

Both equations above are derived from the Kullback-Leibler divergence
(citation here?) and the details are provided in the Appendix.

@@

For non-linearity patterns, the power curve of RESET test climbs
aggressively from 7\% to 85\% as \(log_e(E)\) increases from 0 to 2.5,
while power of other tests respond inactively to the change of effect,
showing that RESET test is much more sensitive to weak non-linear
structure. Meanwhile, no noticeable residual departures can be spotted
from the example residual plots.

>>>>

Effect size in statistics measures the strength of the signal relative
to the noise. It is surprisingly difficult to quantify in general, even
for simulated data as used in this experiment.

For the non-linearity model, the key items defining effect size are
sample size (\(n\)) and variance of the error term (\(\sigma^2\)), and
so effect size would be roughly calculated as
\(\frac{\sqrt(n)}{\sigma}\). As sample size increases the effect size
would increase, but as variance increases the effect size decreases.
However, it is not clear how the additional parameter for the model
polynomial order, \(k\), should be incorporated. Intuitively, the large
\(k\) means more complex pattern, which likely means effect size would
decrease. For the purposes of our calculations we have chosen to use an
approach based on Kullback-Leibler divergence
\citep{kullback1951information}, coupled with simulation. This
formulation defines effect size to be:

\[E = \frac{1}{2}\left(\boldsymbol{\mu}_z'(diag(\boldsymbol{R}_a\sigma^2))^{-1}\boldsymbol{\mu}_z\right)\]

\noindent where
\(\boldsymbol{\mu}_z = \boldsymbol{R}_a\boldsymbol{Z}\boldsymbol{\beta}_z\)
is the expected values of residuals with \(\boldsymbol{Z}\),
\(\boldsymbol{R}_a = \boldsymbol{I}_n - \boldsymbol{H}_a\) is the
residual operator,
\(\boldsymbol{H}_a = \boldsymbol{X}_a(\boldsymbol{X}_a'\boldsymbol{X}_a)^{-1}\boldsymbol{X}_a'\)
is the hat matrix. For the non-linearity model this can be rewritten as

\[E = \frac{1}{2}\left(log\frac{|diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')|}{|diag(\boldsymbol{R}_a)|} - n + tr(diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')^{-1}diag(\boldsymbol{R}_a))\right)\]

\noindent where \(\boldsymbol{V}\) is the covariance matrix of the error
term.

In the heteroskedasticity model, the key elements for measuring effect
size are sample size, \(n\), and the ratio of the biggest variance to
smallest variance, \(b\). Larger values of both would produce higher
effect size. However, it is not clear how to incorporate the additional
shape parameter, \(a\). Thus the same approach is used here, where the
formula can be written as:

\[E = \frac{1}{2}\left(log\frac{|diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')|}{|diag(\boldsymbol{R}_a)|} - n + tr(diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')^{-1}diag(\boldsymbol{R}_a))\right)\]
\noindent where \(diag(.)\) is the diagonal matrix constructed from the
diagonal elements of a matrix.

Derivations for these equations are provided in the Appendix.

To compute the effect size for each lineup we simulate ??

@@

For non-linearity patterns, the power curve of RESET test climbs
aggressively from 16\% to 100\% as \(log_e(E)\) increases from 0 to 2.5,
while power of other tests respond inactively to the change of effect,
showing that RESET test is much more sensitive to weak non-linear
structure. Meanwhile, no noticeable residual departures can be spotted
from the example residual plots.
